---
title: "Homework assignment 2"
subtitle: "Data Science and Machine Learning 1 - CEU 2018"
author: "Peter Paziczki"
date: '2018 február 5'
output: html_document
---

<style>
body {
text-align: justify}
</style>

```{r setup, include=FALSE}
rm(list=ls())
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(data.table)
library(caret)
library(bit64)
library(ggplot2)
library(GGally)
library(gplots)
library(ROCR) # performance function
#library(e1071)
#library(glmnet)
```

# 1. Predicting mental health problems in the tech sector

The variable to predict is treatment.

```{r}
data <- fread("../data/mental-health-in-tech/survey_cleaned.csv")

data <- data[ ,c("comments", "state","work_interfere") := NULL]
data[, age := as.numeric(age)]
data[ , treatment := factor(treatment, levels = c("Yes", "No"))]

## some additional data cleaning
data <- data[age %between% c(18, 100)]

## making treatment variable into a binary variable
data[, treatment_binary := ifelse(treatment == "Yes", 1, 0)]
data <- data[ ,c("Timestamp") := NULL]

## creating large_countries to filter small countries out of the dataset
data[,.N, by = country]
large_countries <- data[, .N, by = "country"][N > 10][["country"]]
data <- data[country %in% large_countries]
```

## 1.1 Explore some predictors that can be used to predict treatment.

I have transformed the `r treatment` variable into a binary variable, denoting 1 to yes. The below chart shows that there is definitely some relation between age and the ratio of people seeking treatment for a mental health condition. It is important to mentione that there is only three observation in the 60-65 range and only one in 70-75 age range.

```{r}
data_by_age <- data[ ,
  .(treatment_rate = mean(treatment_binary), num_obs = .N),
  keyby = .(age_category = cut(age, breaks = seq(0, 100, by = 5), include.lowest = TRUE))]

ggplot(data = data_by_age, aes(x = age_category, y = treatment_rate, size = num_obs)) +
  geom_point() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

The below chart assesses the treatment ratio by gender. The survey has been mostly filled by male employees who, on average, tend to least seek treatment compared to female and trans gender cooworkers. 

```{r}
data[, .N, by = gender]

data_by_gender <- data[ ,
  .(treatment_rate = mean(treatment_binary), num_obs = .N),
  keyby = gender]

ggplot(data = data_by_gender, aes(x = gender, y = treatment_rate, size = num_obs)) +
  geom_point() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

I found it to be interesting to assess how the treatment ratio varies across countries. Having all the countries does not make sense, there are several countries with 1 or a bit more observations.

```{r}
data[,.N, by = country]

data_by_country <- data[ ,
  .(treatment_rate = mean(treatment_binary), num_obs = .N),
  keyby = country]

ggplot(data = data_by_country, aes(x = country, y = treatment_rate, size = num_obs)) +
  geom_point() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

I have dropped the countries that had less than `r 10` observations. Now with still having the 91.5% of the observations, the countries left in the data set could be good predictors. In France and India only a low portion of employees sought treatment, while in Australia it is the highest.

```{r}
large_countries <- data_by_country[num_obs >= 10]

ggplot(data = large_countries, aes(x = country, y = treatment_rate, size = num_obs)) +
  geom_point() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

Those who have a family history of mental illness tend to seek treatment in a much higher ratio than those who don't have. I am expecting it to be a good predictor.

```{r}
data_family_history <- data[ ,
  .(treatment_rate = mean(treatment_binary), num_obs = .N),
  keyby = family_history]

ggplot(data = data_family_history, aes(x = family_history, y = treatment_rate, size = num_obs)) +
  geom_point() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

Those whose employer provides mental health benefits tend to seek treatment in a higher ratio, it also seems to be a good predictor.

```{r}
data_benefits <- data[ ,
  .(treatment_rate = mean(treatment_binary), num_obs = .N),
  keyby = benefits]

ggplot(data = data_benefits, aes(x = benefits, y = treatment_rate, size = num_obs)) +
  geom_point() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

It is quite interesting, but based on the data more difficult is to take advantage of a mental health or substance abuse treatment resources, the more employees tend to seek treatment.

```{r}
data_leave <- data[ ,
  .(treatment_rate = mean(treatment_binary), num_obs = .N),
  keyby = leave]

ggplot(data = data_leave, aes(x = leave, y = treatment_rate, size = num_obs)) +
  geom_point() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

Please find a few examples below where I was expecting to find good predictors but they were found to be not good predictors.

Being self employed does not seem to be a good predictor, there is only a slight difference between the groups in treatment ratio.

```{r}
data_self_employed <- data[ ,
  .(treatment_rate = mean(treatment_binary), num_obs = .N),
  keyby = self_employed]

ggplot(data = data_self_employed, aes(x = self_employed, y = treatment_rate, size = num_obs)) +
  geom_point() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

There is no significant difference in treatment ratio among different company sizes, considering the company size does not seem to be a good predictor.

```{r}
data_no_employee <- data[ ,
  .(treatment_rate = mean(treatment_binary), num_obs = .N),
  keyby = no_employees]

ggplot(data = data_no_employee, aes(x = no_employees, y = treatment_rate, size = num_obs)) +
  geom_point() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

Having the anonymity protected does not seem to play a significant role in seeking or not seeking treatment.

```{r}
data_anonymity <- data[ ,
  .(treatment_rate = mean(treatment_binary), num_obs = .N),
  keyby = anonymity]

ggplot(data = data_anonymity, aes(x = anonymity, y = treatment_rate, size = num_obs)) +
  geom_point() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

## 1.2 Partition your data to 70% training and 30% test samples.

```{r}
set.seed(1234)
training_ratio <- 0.7
train_indices <- createDataPartition(y = data[["treatment"]],
                                     times = 1,
                                     p = training_ratio,
                                     list = FALSE)
data_train <- data[train_indices, ]
data_test <- data[-train_indices, ]
```

## 1.3 Build models with `r glmnet` and `r rpart` that predict the binary outcome of `r treatment` (you don’t have to use all variables if you don’t want to - experiment! Just use the same variables for both model families). Use cross-validation on the training set and use AUC as a selection measure (use `r metric = "ROC"` in train and also don’t forget to use `r classProbs = TRUE, summaryFunction = twoClassSummary` in  trainControl). Make sure to set the same seed before each call to train.

```{r}
# glmnet model for cross-validation, hold-out set AUCs
train_control <- trainControl(method = "cv",
                              number = 10,
                              classProbs = TRUE,
                              verboseIter = TRUE,
                              summaryFunction = twoClassSummary)

tune_grid <- expand.grid("alpha" = seq(0.1 , 1, 0.1),
                         "lambda" = c(0.1, 0.01, 0.001, 0.0001))

set.seed(1234)
glmnet_model <- train(treatment ~
                      poly(age, 2) + gender + country + family_history + benefits + leave,
                      data = data_train,
                      method = "glmnet",
                      preProcess = c("center", "scale"),
                      trControl = train_control,
                      tuneGrid = tune_grid,
                      metric = "ROC")
glmnet_model
```



```{r}
tune_grid <- data.frame("cp" = c(0.1, 0.01, 0.001, 0.0001, 0.00001, 0.000001))

set.seed(1234)
rpart_model <- train(treatment ~ 
                      poly(age, 2) + gender + country + family_history + benefits + leave,
                      data = data_train,
                      method = "rpart",
                      preProcess = c("center", "scale"),
                      trControl = train_control,
                      tuneGrid = tune_grid,
                      metric = "ROC")
rpart_model
```

## 1.4 Compare models based on their predictive performance based on the cross-validation information (you can just use the mean AUC to select the best model).

## 1.5 Evaluate the best model on the test set: draw an ROC curve and calculate and interpret the AUC.

```{r}
# finally, evaluate the best model on test set
test_prediction_glmnet <- predict.train(glmnet_model, 
                                        newdata = data_test, 
                                        type = "prob")
```

Drawing the ROC curve in two different ways

```{r}
glmnet_prediction <- prediction(test_prediction_glmnet$Yes,
                              data_test[["treatment"]])

# built-in plot method
plot(performance(glmnet_prediction, "tpr", "fpr"), colorize=TRUE) 
```

```{r}
# calculate AUC = area under the curve -> the measure of predictive power
AUC <- performance(glmnet_prediction, "auc")@y.values[[1]]
print(AUC)
```

```{r}
# a ggplot version
# using prediction function from ROCR package
glm_perf <- performance(glmnet_prediction, measure = "tpr", x.measure = "fpr")

glm_roc_df <- data.table(
  model = "glm",
  FPR = glm_perf@x.values[[1]],
  TPR = glm_perf@y.values[[1]],
  cutoff = glm_perf@alpha.values[[1]] # cutoff is the treshold
)

ggplot(glm_roc_df) +
  geom_line(aes(FPR, TPR, color = cutoff), size = 2) +
  geom_ribbon(aes(FPR, ymin = 0, ymax = TPR), fill = "blue", alpha = 0.1) +
  geom_abline(intercept = 0, slope = 1,  linetype = "dotted", col = "black") +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, .1)) +
  scale_x_continuous(limits = c(0, 1), breaks = seq(0, 1, .1)) +
  xlab("False Positive Rate") + ylab("True Positive Rate") 
```

## 1.6 If you have to choose a probability threshold to predict the outcome, what would you choose? At this threshold, how large are the true positive rate and the false positive rate? How many false positives and false negatives there are in the test sample?

# 2. Transformed scores

Take the medical appointment no-show dataset we used in class and apply all the cleaning steps we did, then create a training and a test set. Estimate a predictive model of your choice for no_show as a target variable. Get predicted scores (probabilities). Then calculate two transformations of the scores: take the square root and the square of the probabilities. These are valid scores as well, they are alse between 0 and 1 so they can be used for classification.

```{r}
data <- fread("../data/medical-appointments-no-show/no-show-data.csv")

# [... apply the cleaning steps we did in class ...]
# [... create train and test sets ... ]

model <- train(...)

prediction <- predict.train(model, newdata = data_test, type = "prob")$Yes # type = "prob" is imprortant, prediction is a vector,
## it gives a data frame with yes and no in two columns
## each row has a sum of one, my prediction is the yes column.
prediction_sqrt <- sqrt(prediction)
prediction_sq <- prediction^2
```

## 2.1 Draw ROC curves for all three scores and calculate the AUC. How do they compare? Is it surprising in light of the interpretation of the AUC?
## 2.2 What is the key, common property of both the square root and the square functions that leads to this finding?
## 2.3 Draw a calibration plot for all three scores separately:
+group people into bins based on predicted scores
+display on a scatterplot the mean of the predicted scores versus the actual share of people surviving
+How do they compare? Which score(s) can be regarded as well-calibrated probabilites?