---
title: "Homework assignment 3"
subtitle: "Data Science and Machine Learning 1 - CEU 2018"
author: "Peter Paziczki"
date: '2018 február 4 '
output: html_document
---

<style>
body {
text-align: justify}
</style>

```{r setup, include=FALSE}
rm(list=ls())
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(data.table)
library(datasets)
library(MASS)
library(ISLR)
library(caret)
```

# 1. PCA for supervised learning
In this problem we are going to analyze the Boston dataset from the MASS package. The goal will be to predict the variable crim which is the crime rate.

```{r}
data <- data.table(Boston)
```

## 1.1 Do a short exploration of data and find possible predictors of the target variable.

## 1.2 Create a training and a test set of 50%.

## 1.3 Use a linear regression to predict `crim` and use 10-fold cross validation to assess the predictive power.

## 1.4 Try to improve the model

Try to improve the model by using PCA for dimensionality reduction. Center and scale your variables and use `pcr` to conduct a search for the optimal number of principal components. Does PCA improve the fit over the simple linear model?

## 1.5 Use penalized linear models

Use penalized linear models for the same task. Make sure to include lasso (`alpha = 0`) to your tune grid. How does the best model compare to that found in d)? Would pre-processing via PCA help this model? (add `pca` to `preProcess`). Why do you think the answer can be expected?

## 1.6 Evaluate your preferred model on the test set.