---
title: "Lab week 11 - h2o"
subtitle: "Data Science and Machine Learning 2 - CEU 2018"
author: "Jeno Pal"
date: '2018-03-13'
output:
  html_document:
    df_print: paged
  html_notebook:
    df_print: paged
---

H2O: state-of-the-art machine learning software that is even
suitable for big datasets. 
It offers very efficient and scalable implementations of 
popular ML algorithms that can 

* run on distributed systems
* utilize multiple cores
* work with GPUs

Models estimated with h2o can be deployed to production environments
through Java objects (see [here](http://docs.h2o.ai/h2o/latest-stable/h2o-docs/productionizing.html)).
Also, h2o scales well compared to other competitor implementations
(see Szilard Pafka's famous benchmarks [here](https://github.com/szilard/benchm-ml)).

In general, best resource to learn is the
[documentation](http://docs.h2o.ai/h2o/latest-stable/h2o-docs/welcome.html) and many
tutorials are available on YouTube.

```{r}
library(data.table)
library(ISLR)
library(ggplot2)

library(h2o)
h2o.init()
# h2o uses the same algorithms that we have learned before, but doing it in a smarter and more efficient way
# it can use the processors of the GPUs, does not do it by default it, but I can set it to do that
# it is very good especially with large datasets, see Szilárd's github page above
# everything is implemented in java, so it runs a java virtual machine
# it has one node only now, because it is running only on my machine
# i can increase the memory

# everything is happenning on my computer now !!!

# http://localhost:54321/flow/index.html
```

```{r}
# stop it via h2o.shutdown()
```

```{r}
data <- data.table(Wage)
skimr::skim(data)
```

```{r}
data[, c("region","wage") := NULL] # we want to predict log wage, so dropping wage
# for data manipulation with h2o is not so great, it is better to manipulate data first with data table or whatever, then
# turning it into a h2o frame
h2o_data <- as.h2o(data) # it is not a data frame, similar, but not the same, it is an h2o frame
str(h2o_data)
```

```{r}
# splitting data into three parts, train, validation and test, it creates lists
splitted_data <- h2o.splitFrame(h2o_data, 
                                ratios = c(0.6, 0.2), # first part 60%, second 20%, third has the rest
                                seed = 123)

nrow(data_train)
nrow(data_valid)
nrow(data_test)

data_train <- splitted_data[[1]]
data_valid <- splitted_data[[2]]
data_test <- splitted_data[[3]]
```

GLM: __penalized__ linear regression methods.
```{r}
y <- "logwage"  # outcome variable's name
X <- setdiff(names(h2o_data), c(y, "year")) # a base R function
# taking all the columns of h2o data except the year column
# it is a character vector with all the features i want to use

glm_fit <- h2o.glm(X, y, # running old logistic regression, it is a penalized model here, x - features, y - outcome
               training_frame = data_train, # the data used for training
               alpha = 0, # alpha can be between 0 (Ridge) and 1 (Lasso), tells is whether it is lasso or ridge or anything in between
               nfolds = 5, # specifying cross-validation (5-fold in this case)
               seed = 123)

# xval stands for 'cross-validation'
print(h2o.performance(glm_fit, xval = TRUE)) # xval = TRUE asks whether it needs to report metrics or not, without it i can ask for training error
# RMSE is the average of the five models estimated via 5-fold cross-validation
```

```{r}
print(h2o.rmse(glm_fit, xval = TRUE)) # giving RMSE coming from the CV above
```

You can perform a grid search with cross-validation to tune hyperparamters.
```{r}
# let's do hyperparameter tuning
hyper_params <- list(alpha = c(0, .25, .5, .75, 1)) # exploring ridge, lasso and models in between

# build grid search with previously selected hyperparameters
glm_grid <- h2o.grid(x = X, # it is a grid search, that is the reason for calling it grid ...
                     y = y, 
                     training_frame = data_train, 
                     algorithm = "glm", # it could be set to any other algorithm
                     lambda_search = TRUE, # penalized linear models have two parameters in caret, alpha and lambda
                                           #here lambda is a way to ask h2o to find a good lamdba value
                     # grid_id = "my_glm_grid" - it is a way to name my grid, but shouldn't do it,
                     # because after naming i won't be able to update it anymore
                     nfolds = 5, # 5-fold CV
                     seed = 123,
                     hyper_params = hyper_params)
h2o.getGrid(grid_id = glm_grid@grid_id, sort_by = "rmse", decreasing = FALSE)
# i need to give which grid to evaluate, but here it is not enough to say glm_grid, i need to refer to the id of the grid
str(glm_grid)

# the grid id:
glm_grid@grid_id

# after referring to the grid_id, I need to tell to sort the results by RMSE in a decreasing fashion
# it ranks the models, 0.25 was the best performing

# http://docs.h2o.ai/h2o/latest-stable/h2o-docs/data-science.html
```

Random Forests are there for you to use as well.
```{r}
# random forests
rf_params <- list(ntrees = c(500), # number of trees to estimate
                  mtries = c(2, 3, 5))

rf_grid <- h2o.grid(x = X, 
                    y = y, 
                    training_frame = data_train, 
                    algorithm = "randomForest", 
                    nfolds = 5,
                    seed = 123,
                    hyper_params = rf_params)

h2o.getGrid(grid_id = rf_grid@grid_id, sort_by = "rmse", decreasing = FALSE)
```

Just like GBMs.

```{r}
# GBM hyperparamters
gbm_params <- list(learn_rate = c(0.01, 0.05), # shrinkage parameter, the lower the value the smaller steps tha algo takes
                    max_depth = c(2, 3, 5), # complexity of the trees
                    sample_rate = c(0.5), # using only half of the data to estimate trees
                    col_sample_rate = c(0.5, 1.0)) # constraining the features to build trees
# Train and validate a cartesian grid of GBMs
gbm_grid <- h2o.grid(x = X, 
                     y = y, 
                     training_frame = data_train, 
                     algorithm = "gbm", 
                     nfolds = 5,
                     seed = 123,
                     ntrees = 300,
                     hyper_params = gbm_params)

h2o.getGrid(gbm_grid@grid_id, sort_by = "rmse", decreasing = FALSE)
```

```{r}
# get best models
glm_model <- h2o.getModel(h2o.getGrid(glm_grid@grid_id)@model_ids[[1]]) # getting the grid id, then getting the model
# need to choose the first model, because models are sorted by RMSE in a decreasing fashion, so the first one is the best performing model
rf_model <- h2o.getModel(h2o.getGrid(rf_grid@grid_id)@model_ids[[1]])
gbm_model <- h2o.getModel(h2o.getGrid(gbm_grid@grid_id)@model_ids[[1]])

# predict on validation set
validation_performances <- list(
  "glm" = h2o.rmse(h2o.performance(glm_model, newdata = data_valid)),
  "rf" = h2o.rmse(h2o.performance(rf_model, newdata = data_valid)),
  "gbm" = h2o.rmse(h2o.performance(gbm_model, newdata = data_valid))
)

validation_performances
```

```{r}
# test set performance
h2o.rmse(h2o.performance(gbm_model, newdata = data_test))
```

```{r}
# turn back h2oFrames to plain data.frame
prediction_vs_truth = data.table(cbind(
    as.data.frame(h2o.predict(gbm_model, newdata = data_test)), # it produces a h2o frame, as.data.frame transforms it back to a data frame
    as.data.frame(data_test)[["logwage"]]
))

setnames(prediction_vs_truth, c("prediction", "truth"))

head(prediction_vs_truth)

ggplot(data = prediction_vs_truth,
       aes(x = truth, y = prediction)) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  xlim(3, 6) + ylim(3, 6) +
  geom_point() +
  theme_minimal()
```

And, of course, there is deep learning, too! We'll come to that later.
Let's just try it out of the box and see validation set performance.
```{r}
dl_model <- h2o.deeplearning(X, y, 
                             training_frame = data_train,
                             reproducible = TRUE, # needed in deeplearning for full reproducibility, setting the seed is not enough ...
                             # it does not utilize everything it could, if setting it to TRUE
                             seed = 123)
h2o.rmse(h2o.performance(dl_model, newdata = data_valid))
```

XGboost is also available (however, not on Windows).
